---
title: "Data manipulation"
output: github_document
date: "2024-09-19"
---

```{r setup, echo=FALSE, message=FALSE}
library(tidyverse)
```

This document will show how to *manipulate* data

Import the two datasets that we need
```{r}
litters_df <- read_csv("data/FAS_litters.csv",
                       na = c("NA", "", "."),
                       show_col_types = F) 

litters_df <- janitor::clean_names(litters_df)

pups_df <- read_csv("data/FAS_pups.csv",
                    na = c("NA","","."),
                    show_col_types = F)

pups_df<- janitor::clean_names(pups_df)
```


## `select`

Use `select` to select variables
```{r}
select(litters_df, group, litter_number, gd0_weight)
```
Select from columns A to B with :
```{r}
select(litters_df, group:gd18_weight)
```

drop a column using -
```{r}
select(litters_df, -pups_survive)
```

combine multiple operations
```{r}
select(litters_df, -(group:gd18_weight))
```

manipulate in more detail
```{r}
select(litters_df, starts_with("gd"))

select(litters_df, contains("pups"))
```

can rename within select, or use `rename`
```{r}
select(litters_df, GROUP = group)

rename(litters_df, GROUP = group)

# change order of columns using select or relocate
select(litters_df, litter_number, gd0_weight, everything())

relocate(litters_df, litter_number, gd0_weight)
```

Learning Assessment: In the pups data, select the columns containing litter number, sex, and PD ears.
```{r}
select(pups_df, litter_number,sex,pd_ears)
```

## `filter`

`select` works on columns, `filter` works on rows
```{r}
filter(litters_df,gd_of_birth == 20)
filter(litters_df,gd_of_birth == 19)
```
different logical operators
```{r}
filter(litters_df, pups_born_alive > 8)
filter(litters_df, pups_born_alive >= 8)
filter(litters_df, pups_born_alive != 9)
```
can use with non-numeric types also
```{r}
filter(litters_df, group == "Low8")

filter(litters_df, group %in% c("Low8","Low7"))
```

```{r}
filter(litters_df, group %in% c("Low7", "Low8"), pups_born_alive == 8)
```

filter to drop na: drop_na
```{r}
drop_na(litters_df)

drop_na(litters_df, gd0_weight)
```

Learning Assessment: In the pups data:

Filter to include only pups with sex 1
Filter to include only pups with PD walk less than 11 and sex 2

```{r}
filter(pups_df, sex == 1)

filter(pups_df, pd_walk < 11, sex == 2)
```

##  `mutate`

```{r}
mutate(litters_df, wt_gain = gd18_weight - gd0_weight)

mutate(litters_df, sq_pups = pups_born_alive^2)

mutate(litters_df, group = str_to_lower(group))

mutate(litters_df, 
  wt_gain = gd18_weight - gd0_weight,
  group = str_to_lower(group))
```

## `arrange`

pretty straightforward
```{r}
arrange(litters_df, gd0_weight)

#descending order
arrange(litters_df, desc(gd0_weight))
arrange(litters_df, -gd0_weight)

arrange(litters_df, pups_born_alive,gd0_weight)
```

## PIPING!
The best thing to happen to R since sliced bread

```{r}
litters_df <- read_csv("data/FAS_litters.csv",
                       na = c("NA","","."),
                       show_col_types = F) |>
  janitor::clean_names() |>
  select(-pups_born_alive) |>
  filter(group == "Con7") |>
  mutate(wt_gain = gd18_weight - gd0_weight)
```

