---
title: "data_import"
output: github_document
date: "2024-09-17"
author: "Ravi Brenner"
---

```{r setup, echo=FALSE, message = FALSE}
library(tidyverse)
library(janitor)
library(readxl)
library(haven)
```


This document will show how to import data into R
following along with in class work

## import FAS litters CSV

```{r}
litters_df <- read_csv("./data/FAS_litters.csv") 

litters_df <- clean_names(litters_df) #from janitor package
```

Look at the dataset briefly
```{r}
head(litters_df)
tail(litters_df)
```

```{r, eval=FALSE}
view(litters_df)
```

## import FAS pups dataset

```{r, message=FALSE}
pups_df <- read_csv("./data/FAS_pups.csv")

pups_df <- clean_names(pups_df)
```

look at the data
```{r}
head(pups_df)
```

## look at read_csv options

col_names and skipping rows

```{r, message=FALSE}
litters_df <- 
  read_csv(
    file = "data/FAS_litters.csv",
    skip = 2
  )
```

missing data
```{r}
litters_df <- 
  read_csv(
    file = "data/FAS_litters.csv",
    skip = 0
  )

litters_df
```
can see gd0_weight and gd18_weight is character which isn't what we want
inspecting the data, we see a mix of blanks, NAs, and . to indicate missing values

can fix this in read_csv
```{r}
litters_df <- 
  read_csv(
    file = "data/FAS_litters.csv",
    na = c("","NA",".")
  )

litters_df <- clean_names(litters_df)
```

A nice function to maybe add to my cleaning workflow
```{r}
skimr::skim(litters_df)
```

what if we code `group` as a factor variable?
```{r}
litters_df <- 
  read_csv(
    file = "data/FAS_litters.csv",
    na = c("","NA","."),
    col_types = cols(Group = col_factor())
  )

litters_df <- clean_names(litters_df)
```

## importing from excel
```{r}
mlb_df <- read_excel("data/mlb11.xlsx",
                     sheet = "mlb11")
```

## importing from SAS
```{r}
pulse_df <- read_sas("data/public_pulse_data.sas7bdat")
```

